Limited plasticity but increased variance in physiological rates across ectotherm populations under climate change

### 2024-07-15 ms.qmd
New abstract merging ideas with Franks suggestions

Climate change is expected to result in warmer and more variable thermal environments globally. Maintaining variability in physiological rates and adjusting them in response to extreme temperatures (plasticity) is essential for allowing populations to adapt to climate change. Yet, incorporating both plasticity and changes in phenotypic variation when predicting the impacts of climate change on populations is often ignored. We developed effect size estimates that quantify how physiological rate variance changes relative to the mean as temperature increases, which we refer to as 'relative variance'. We show by meta-analysing >1900 effects from 226 species that relative variance in physiological rates decreases at higher temperatures driven by the lower proportional increase in physiological rate variance relative to the mean. Freshwater ectotherms are capable of acclimating and have the smallest reductions in relative variance. Marine organisms also acclimated, but plasticity traded-off with a reduction in relative variance in physiological rates at higher temperatures. Relative variance reductions were particularly pronounced for terrestrial ectotherms, and this coincided with a lack of capacity for acclimation, highlighting the vulnerability of terrestrial ectotherms to climate change. Our results show that, while there are weak acclimation responses in some habitats (marine and freshwater), such responses may trade-off with reductions in relative physiological rate variance that may have important evolutionary and ecological ramifications that affect our understanding of how climate change will impact populations now and in the future.

@@

Climate change is expected to result in warmer but also more variable thermal environments globally [@Suarez2023; @Easterling2000;
@Ummenhofer2017]. Greater thermal variability is predicted to result in strong selection pressures that lead to genetic adaptation and/or the evolution of adaptive phenotypic plasticity -- both of which are considered important for population resilience to contemporary human-induced climate change [@Seebacher2015-wi; @Merila2014; @Chevin2015; @Chevin2017; @Chevin2010; @Nunney2016; @seebacher2023can; @cooke2021one]. Without plasticity or adaptation, high extinction rates are expected unless organisms can migrate to track suitable habitats [@Cahill2012; @Nunney2016].

>>>>

Climate change will result in warmer and more variable temperatures globally impacting physiological rates and function in ectothermic animals. Acclimation of physiological rates can help maintain function, however, how variance in physiological rates change with temperature has been ignored despite its potential ecological and evolutionary importance. We apply new effect sizes that quantify the extent to which both mean and variance in physiological rates change as temperature increases. Meta-analysing >1900 effects from 226 species, we show that variance in physiological rates increases at higher temperatures, but that the magnitude of change depends on habitat. Freshwater and marine ectotherms are capable of acclimation and have the greatest increase in physiological rate variance. In contrast, terrestrial ectotherms lack a capacity for acclimation and have a smaller increase in variance. Simulations suggest fundamental differences in among-individual variation in thermal breadth and optima of performance curves across habitats that contribute to these differences. Our results highlight the greater vulnerability of terrestrial ectotherms to climate change because limited increases in physiological rates will provide less raw material for evolutionary adaptation. Considering both acclimation capacity and variance in physiological rates side-by-side is therefore important for understanding how climate change will impact populations.

@@

Climate change is expected to result in warmer and more variable thermal environments globally [@Suarez2023; @Easterling2000; @Ummenhofer2017]. Greater thermal variability is predicted to result in strong selection pressures that lead to genetic adaptation and/or the evolution of adaptive phenotypic plasticity -- both of which are considered important for population resilience to contemporary human-induced climate change [@Seebacher2015-wi; @Merila2014; @Chevin2015; @Chevin2017; @Chevin2010; @Nunney2016; @seebacher2023can; @cooke2021one]. Without plasticity or adaptation, high extinction rates are expected unless organisms can migrate to track suitable habitats [@Cahill2012; @Nunney2016].
